---
title: "Group 20"
format: 
  html:
    embed-resources: true
    code-tools: true
  pdf: default
editor_options: 
  chunk_output_type: console
execute:
  echo: false
  eval: true
  warning: false
  message: false
---

```{r}
library(dplyr)
library(ggplot2)
library(faraway)
library(gridExtra)
library(MASS)
library(plyr)
library(dplyr)
library(sjPlot)
```

# Introduction

The main objective is to know **which factors influence the number of days an animal spends in the shelter before their final outcome is decided**.

# Explanatory Data

## Data Distribution

**The response variable**:

-   Time_at_Shelter -- Days spent at the shelter between being admitted and the final outcome.

**The response variable**:

-   The Animal_type -- The type of animal admitted to the shelter

-   Month -- Month the animal was admitted, recorded numerically with January=1

-   Year. -- Year the animal was admitted to the shelter.

-   Intake_type -- Reason for the animal being admitted to the shelter

-   Chip_Status -- Did the animal have a microchip with owner information?

```{r}
library(ggplot2)
library(dplyr)
library(ggbeeswarm)
library(car)
```

```{r}
#| echo: false
#read data
data <- read.csv('dataset20.csv', stringsAsFactors = FALSE)

# Data cleaning
# Check for missing values
sum(is.na(data))
# Remove rows with missing values
data <- na.omit(data)
write.csv(data, 'data.csv')
```

```{r}
# If 'month' and 'year' are used to create dates, it makes sense
data$date <- as.Date(paste(data$year, data$month, "01", sep = "-"))

# Convert 'animal_type', 'intake_type', 'output_type', 'chip_status' to factors
data$animal_type <- factor(data$animal_type)
data$intake_type <- factor(data$intake_type)
data$outcome_type <- factor(data$outcome_type)
data$chip_status <- factor(data$chip_status)
```

```{r}
# the distribution of animal time in shelters
ggplot(data, aes(x=time_at_shelter)) + 
  geom_histogram(binwidth=1, fill='blue', color='black')
```

Most animals stay in the shelter for a short period of time, while a few animals stay in the shelter for a long time. Although the histogram itself does not display outliers, the long tail suggests that there are relatively few cases that have stayed in the shelter for a very long time, which may be outliers.

## Relationship between Variables

```{r}
# the distribution of animal time in shelters
ggplot(data, aes(x=time_at_shelter)) + 
  geom_histogram(binwidth=1, fill='blue', color='black')
```

## Data Visualization

```{r}
# Data visualization
# Box plot of animal types and shelter time
ggplot(data, aes(x=animal_type, y=time_at_shelter, fill=animal_type)) + 
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(palette="Set1")

# Box plot of whether to implant chips and shelter time
ggplot(data, aes(x=chip_status, y=time_at_shelter, fill=chip_status)) +
  geom_boxplot() +
  theme_minimal() +
  scale_fill_brewer(palette="Set1")

# Box plot of animal types and reasons for containment
ggplot(data, aes(x=intake_type, y=time_at_shelter, fill=animal_type)) + 
  geom_boxplot() +
  facet_wrap(~animal_type) + # Use face_wrap to divide faces by animal type
  theme_minimal() +
  scale_fill_brewer(palette="Set1") +
  labs(title="Time at Shelter by Intake Type and Animal Type",
       x="Intake Type",
       y="Time at Shelter (days)")

# Box plot of animal types and the outcome of leaving the shelter
ggplot(data, aes(x=outcome_type, y=time_at_shelter, fill=animal_type)) + 
  geom_boxplot() +
  facet_wrap(~animal_type) +
  theme_minimal() +
  scale_fill_brewer(palette="Set1") +
  labs(title="Time at Shelter by Outcome Type and Animal Type",
       x="Outcome Type",
       y="Time at Shelter (days)")

# Bar chart with chip status and containment reasons
ggplot(data, aes(x=intake_type, fill=chip_status)) + 
  geom_bar(position="dodge") +
  facet_wrap(~chip_status) +
  theme_minimal() +
  scale_fill_brewer(palette="Set1") +
  labs(title="Intake Reasons by Chip Status",
       x="Intake Type",
       y="Count")

# Bar chart with chip status and outcome of leaving the shelter
ggplot(data, aes(x=outcome_type, fill=chip_status)) + 
  geom_bar(position="dodge") +
  facet_wrap(~chip_status) +
  theme_minimal() +
  scale_fill_brewer(palette="Set1") +
  labs(title="Outcome Types by Chip Status",
       x="Outcome Type",
       y="Count")
```

-   Box plot of animal types and shelter time:

The red box represents animals with chips. The median residence time of these animals is low, indicating that when animals have chips, they usually stay in shelters for a shorter period of time. The blue box represents animals without chips. This category of animals has a longer residence time in the shelter, with a higher median and quartile range than animals with chips, indicating significant fluctuations in residence time. The green box may indicate animals that are unable to scan chips for certain reasons. There are very few data points in this group, making it difficult to draw specific conclusions.

-   Box plot of whether to implant chips and shelter time:

The shorter box of birds indicates that their stay time in the shelter is relatively consistent and short. The height of the cat's enclosure is moderate, indicating a certain fluctuation in their stay time in the shelter. The box of dogs shows a wide range of stay times, with a median higher than that of cats and birds, and some high outliers indicating that some dogs have stayed in the shelter for a very long time. The box height and median of wild animals are similar to those of cats, but also contain some higher outliers.

-   Box plot of animal types and reasons for containment:

From the average residence time of animals that have not been received from different animals, we can see that there are more data points for cats and dogs, and cats have a shorter average residence time when confiscated and actively sent out, while dogs have a longer residence time when wandering. The sample size of birds and wildlife is relatively small, and the data distribution is also relatively scattered.

-   Box plot of animal types and the outcome of leaving the shelter:

Regardless of the type of outcome, there is a certain diffusion in the residence time of cats and dogs, indicating diversity in their final outcomes. Some dogs spend a very long time in shelters before being euthanized, which may reflect the challenge of finding a home or dealing with health issues for these dogs.

-   Bar chart with chip status and containment reasons:

Most chipless animals are sent to shelters due to wandering, which may indicate that more chipless animals are wandering outside. There is a certain relationship between the chip scanning results and the reasons why animals are being housed, for example, animals with chips are often voluntarily sent out by their owners.

-   Bar chart with chip status and outcome of leaving the shelter:

Most chipless animals are ultimately adopted, possibly because many of them were initially brought into shelters as stray animals. Animals with chips are more likely to be retrieved by their original owners

Due to the fact that in a scatter plot, when there are too many data points, they may overlap, making it difficult to observe all the data points. We use a combination of bee plot and violin plot. The bee plot disperses points along a certain width of the classification axis, making each point clearly visible, while the violin plot shows the density of data distribution

```{r}
# Violin chart combined with bee colony chart, showing the impact of animal types on the time spent in shelters
ggplot(data, aes(x=animal_type, y=time_at_shelter, color=animal_type)) + 
  geom_violin(trim=FALSE) +
  geom_beeswarm() +
  theme_minimal() +
  labs(title="Time at Shelter by Animal Type",
       x="Animal Type",
       y="Time at Shelter (days)")

# Violin chart combined with bee colony chart to demonstrate the impact of chip status on the time spent in the shelter
ggplot(data, aes(x=chip_status, y=time_at_shelter, color=chip_status)) + 
  geom_violin(trim=FALSE) +
  geom_beeswarm() +
  theme_minimal() +
  labs(title="Time at Shelter by Chip Status",
       x="Chip Status",
       y="Time at Shelter (days)")
```

-   Charts classified by animal type:

The distribution of each animal type has different shapes and widths, indicating a different distribution of their residence time in shelters. The distribution of dog stay time is relatively wider, with several abnormally high data points, which may indicate that some dogs require longer time to have final results. The data distribution of cats is relatively concentrated, but there are also several higher data points. There are relatively few data points for birds and wildlife, which may make it difficult to draw statistically significant conclusions.

-   Chart classified by chip status:

Animals without chips exhibit significant variability in their residence time in shelters, including some extreme values. The residence time of animals with chips seems to be relatively short. In the category that cannot scan chips, animals have a very short stay in the shelter, but there are very few data points in this category.

## Summary Statistics

```{r}
# Summary statistics
summary_stats <- summary(data)

print(summary_stats)
```

-   Animal_type:

There are four types of animals, with the highest number of dogs (1146), 304 cats, 3 birds, and 12 wild animals.

-   Month:

represents the month in the dataset, with a minimum value of 1 (January) and a maximum value of 12 (December). The median is 6.

-   Year:

The dataset contains data from 2016 and 2017.

-   Intake_type:

Describe the reasons why animals enter a shelter, mainly including being confiscated, being sent out for adoption by the owner, and wandering.

-   Output_type:

Describe the results of an animal leaving a shelter, including adoption, death, euthanasia, foster care, and return to its owner.

-   Chip_status:

Record whether animals have chips. There are 278 animals with chips, 1127 animals without chips, and 60 animals that cannot be scanned with chips.

-   Time_shelter:

The amount of time an animal stays in a shelter, from 0 days to a maximum of 87 days. The average stay time is about 6 days, with a median of 4 days.

-   Date:

The range of date data is from October 1, 2016 to September 1, 2017. The median date is April 1, 2017.

## Correlation Analysis

```{r}
# Select only numerical variables
numeric_data <- data[, c("month", "year", "time_at_shelter")]

# Using the cor() function to calculate the correlation coefficient matrix
correlation_matrix <- cor(numeric_data, use = "complete.obs") # Processing missing values for the 'use' parameter

print(correlation_matrix)
```

There is a strong negative correlation (-0.74) between month and year, which means that as the year increases, the month value decreases.

The correlation between month and time_shelter is almost zero (-0.008), indicating that there is no significant linear relationship between the month and the time the animal spends in the shelter.

The correlation between year and time_shelter is also very low (-0.039), indicating that there is no significant linear relationship between year and the time the animal stays in the shelter.

## Identifying Outliers
```{r}
# Calculate IQR
Q1 <- quantile(data$time_at_shelter, 0.25)
Q3 <- quantile(data$time_at_shelter, 0.75)
IQR <- Q3 - Q1

# Define the upper threshold for outliers
upper_bound <- Q3 + 1.5 * IQR

# Replace outliers with median
data$time_at_shelter[data$time_at_shelter> upper_bound] <- median(data$time_at_shelter)

# animal_type
prop.table(table(data$animal_type))
# think twice
data <- data %>%
  filter(!(animal_type=="BIRD" | animal_type=="WILDLIFE"))
```

## Multicollinearity
```{r}
# Fit a linear model with 'time_t_shell' as the response variable and all other columns as the predictive variables
model <- lm(time_at_shelter ~ ., data = data)
# Calculate and print VIF values
vif_values <- vif(model)
print(vif_values)
```
The GVIF values for month, year, and date are very high, especially for year and date, with adjusted GVIF values of approximately 241.73 and 163.95, respectively, far exceeding the conventional critical values, indicating a high degree of linear relationship between them and other variables in the model.

The adjusted GVIF values for animal-type, intake_type, output_type, and chip_status are relatively low, indicating that their collinearity with other variables is relatively small.
